Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-22 index.qmd
section:issues, aseV; process:write

<!-- 3. Example-->
These concepts are more easily understood through an example. For instance, in the context of evaluating text quality, @fig-discriminal_process could depict the underlying discriminal process distributions for two written texts, highlighting differences in their discriminal dispersions and modal discriminal processes along the quality trait continuum. Furthermore, @fig-discriminal_difference could display the discriminal difference distribution for these texts, showing that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area. Consequently, the dichotomous outcome of this comparison would favor text A.

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (see @fig-logistic_vs_normal).

@@

<!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
The assumption of *zero correlation between stimuli* is more easily understood through an example. For instance, when using pairwise comparisons to evaluate text quality, the assumption suggests that the judge's perception of the trait in one text does not influence his perception of the same trait on another. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. This cancellation was mathematically demonstrated by @Andrich_1978, assuming discriminal processes with additive biases. However, it is easy to imagine two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 4. The first scenario where the assumption does not hold -->
In the first scenario, evidence suggests that when judges are confronted with multi-dimensional, complex traits and heterogeneous stimuli, such as when evaluating written text quality, their assessments are influenced by multiple, intricate aspects of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. In such cases, it is not inconceivable that various aspects of the stimuli unevenly influence the judges’ perception of one text over another, leading to biases that resist cancellation, as the effects of the factors involved in the comparison are neither equally weighted nor opposing. For example, this may occur when a judge, while assessing the argumentative quality of a text, places disproportionate emphasis on grammatical errors, ultimately favoring texts with fewer errors but weaker arguments. Although the literature does not provide direct evidence for this specific argument, studies such as @Pollitt_et_al_2003 do offer evidence of judges' biases, supporting the idea that the factors involved in the pairwise comparisons may not always cancel out.

@@

<!-- 5. The second scenario where the assumption does not hold -->
In the second scenario, although the CJ literature has acknowledged the existence of additional hierarchical structures relevant to the stimuli, the statistical handling of this extra source of correlation has been inadequate. For example, in cases where data included multiple samples of stimuli within different individuals, researchers have often relied on using (averaged) estimated BTL scores to conduct further analyses and tests at the hierarchical level of the individuals [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach has the notable limitation of ignoring the uncertainty associated with the scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).

@@

<!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
In contrast, the assumption of *equal dispersion between stimuli* suggests that the perceived trait of all texts varies around their modal discriminal process in the same way across comparisons. While Thurstone noted that the assumption of equal dispersion may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp. 374], no study explicitly proposes that this assumption may also be violated due to the presence of an additional hierarchical structure relevant to the texts. One such scenario could occur when comparing university and secondary school students. In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students might exhibit a broader range of writing abilities, leading to greater variability in their text quality. Despite the contrived nature of the example, it helps to effectively illustrate how assuming equal dispersions in the texts can overlook meaningful differences in the reliability of text quality across groups or individuals.

@@

<!-- X. But what are the problems when the zero correlation assumption does not hold? -->

@@

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that the BTL model commonly functions as the measurement model for the trait of interest in CJ experiments [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
However, the statistical literature advises caution when using estimated scores for additional analyses and tests, as well as when eliminating data through ad hoc univariate procedures. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020]. Finally, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of type-I or type-II errors [@McElreath_2020].

@@

## The assessment design features and their role on reliability and validity {#sec-theory-issue3}

>>>>

<!-- 3. Example-->
These concepts are more easily understood through an example. For instance, in the context of evaluating text quality, @fig-discriminal_process could depict the underlying discriminal process distributions for two written texts, highlighting differences in their discriminal dispersions and modal discriminal processes along the quality trait continuum. Furthermore, @fig-discriminal_difference could display the discriminal difference distribution for these texts, showing that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area. Consequently, the dichotomous outcome of this comparison would probably favor text A.

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (see @fig-logistic_vs_normal).

@@

<!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
The assumption of *zero correlation between stimuli* can be better understood through an example. For instance, when using pairwise comparisons to evaluate text quality, the assumption implies that a judge's perception of a trait in one text does not influence his perception of the same trait in another text. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. This cancellation was mathematically demonstrated by @Andrich_1978, using the BTL model under the assumption of additive biases in the discriminal processes. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 4. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits and heterogeneous stimuli may introduce dependencies between stimuli. Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. In this context, it is not inconceivable that these aspects, being neither equally weighted nor opposing, may unevenly influence judges' perceptions, resulting in biases that resist cancellation. For example, this might occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of judges' biases, supporting the idea that the different factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 5. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by the additional hierarchical structure may introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Nevertheless, despite recognizing such hierarchical structures in CJ data, the statistical handling of this extra source of dependency in the CJ literature has been inadequate. For instance, in cases where the CJ data included multiple samples of stimuli from the same individuals, researchers have often relied on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. This approach, however, has the significant limitation of ignoring the uncertainty associated with the scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).

@@

<!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimuli is the same across stimuli. While Thurstone acknowledged that this assumption may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp. 374], no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts. One such scenario might arise, for example, when comparing texts produced by university and secondary school students. In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts. Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.

@@

<!-- 7. But what are the challenges when these assumptions are violated? -->
["however" paragraph]{style="color:red;"}

@@

<!-- 8. So, how do we mitigate these risks? -->
["to mitigate" paragraph]{style="color:red;"}

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that the BTL model commonly functions as the trait's measurement model in CJ experiments [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
However, the statistical literature advises caution when using estimated scores for additional analyses and tests, as well as when eliminating data through ad hoc univariate procedures. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020]. Finally, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].

@@

## The diverse assessment design features and their role on reliability and validity {#sec-theory-issue3}
